---
title: "Exploratory Data Analysis"
author: "Kaleb Combs kec2746"
date: "10-19-2019"
output: html_document
---

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(fivethirtyeight)
```

The two datasets I chose for this project are US_births_2000_2014 and storms. The US_births_2000_2014 dataset provides detailed information about the total number of babies born in the U.S. from 2000-2014. Considering that many of the births in the U.S. are scheduled (either through induced delivery or c-section), this dataset attempts to look at the impact of superstition on due dates. The second dataset, storms, tracked 198 U.S. tropical storms ranging from year 1975 to 2014. 'Storms' gives detailed information about the storm name, date of report, status (tropical, hurricane, etc.), category (ranging from -1 to 5), wind speed (knots), storm pressure (millibars), and diameter of the impacted area based on whether tropical or hurricane strength winds were experienced. Considering the profound impact that weather has on our everyday lives, I chose these datasets in an attempt to find possible relationships between storm intensity and birth frequency from 2000 to 2014.

```{r}
head(US_births_2000_2014)
head(storms)
```

These datasets have two variables in common: month and year. The range for month in both datasets is 1-12. The year variable in the storms dataset starts in 1975 while the year variable for the US_births_2000_2014 dataset starts in 2000. I did an inner join so that only the rows with matching obervations in both datasets from year 2000-2014 would be included. All of the cases recorded before the year 2000 for the storm dataset were dropped. This can be problematic because it will limit the amount of observable change for the storms dataset. For instance, if all of the years leading up to 2000 had a much higher frequency of category 4 and 5 storms, then the numeric data for these years would be noticeably different. All of the other information from both datasets was retained. 

```{r}
inner_join(US_births_2000_2014, storms, by="year") ->Joined
glimpse(Joined)
```

Although this did join the two datasets on the basis of year and only included observations from 2000-2014, there's still a lot of extraneous information that I do not plan on using to investigate the relationship between storm intensity and birth frequency. 

```{r}
Joined%>%select(-day, -date, -hour, -lat, -long, -day_of_week, -date_of_month)->Revised
head(Revised)
```

In order to gain a better understanding of the information in 'Revised', I'll use common dplyr functions to explore and manipulate the dataset.

Since we're interested in looking at whether or not our tropical storms have any impact on birth frequency, let's begin by creating a variable that categorizes storms based on perceived strength. Below, a new variable called 'storm_cat' is created to categorize storm intensity. 

```{r}
Revised%>%mutate(storm_cat = case_when(category>3~"strong", 2<=category&category<=3~"medium", category<2~"weak"))->Revised2
head(Revised2)
```

Great! Now let's start exploring all of the information that this dataset has to offer. Hurricanes are really interesting to me so I wanted to figure out which hurricane over the fourteen year period had the highest average wind speed.

```{r}
head(Revised2%>%filter(status=="hurricane")%>%group_by(year, name)%>%summarize(mean_wind=mean(wind))%>%arrange(desc(mean_wind)))
```

Hurricane Ivan in 2004 had the highest average wind speed for all hurricanes recorded in this dataset.

But is this the fastest wind speed recorded over the fourteen year period?
```{r}
glimpse(Revised2%>%filter(status=="hurricane")%>%group_by(name, year)%>%summarize(max_wind=max(wind))%>%arrange(desc(max_wind)))
```

Although hurricane Ivan is close to the maximum recorded wind speed as well, we can see that hurricane Wilma in 2005 is the fastest with wind speeds of 160 knots.

If we refer back to our joined/modified dataset, 'Revised2', we can see that there are two variables month.x and month.y that both convey similar information. Let's rename them to avoid confusion.

```{r}
Revised2%>%rename(Birth_Month=month.x, Storm_Month=month.y)->Revised3
head(Revised3)

```

We know from the information provided that there are 198 storms included in this dataset from 1975-2014, but we removed a large portion of these when joining by a common year. Given our new dataset, Revised3, how many distinct storms were there from 2000-2014 and in which year did they occur? 

```{r}
head(Revised3%>%group_by(year)%>%summarize(n_storms=n_distinct(name), mean_births=mean(births)))

```

What if we wanted to know how many of them belonged to each category and in which year they occurred?

```{r}
head(Revised3%>%group_by(year, category)%>%summarize(n_distinct=n_distinct(name)))
```

From the first six rows, we can see that for the year 2000, the number of distinct storms already sums to a number greater that the total number of unique storms recorded for that year when grouped by year alone. This suggests that multiple storms can be found within each category. As a storm gains strength, it retains the same name, but transitions into a new category. Thus, it's recognized as a unique entry for multiple categories, but only as one unique entry for the year it belongs to.

When trying to assess the effects of storm intensity, it might be helpful to find out which year had the greatest number of category 5 hurricanes. 

```{r}
head(Revised3%>%filter(category=="5")%>%group_by(year, name)%>%summarize(n_5=n_distinct(name)))
```

It looks like 2005 was a very devastating year with a total of four category 5 hurricanes.

Let's introduce a few more columns. 'Revised3' currently lists wind speed in knots. Let's create a variable for wind speed in miles per hour (mph), kilometers per hour (kph), and then take the inverse of wind speed in mph and call it inverse_mph.

```{r}
Revised3%>%mutate(wind_mph=wind*1.151)%>%mutate(wind_khp=wind_mph*.62137)%>%mutate(mph_inverse=1/wind_mph)->Final
head(Final)

```

Now that the dataset has been modified and explored, we can use summary statistics to learn more about specific variables. 

```{r}
Final%>%select(births, wind, pressure)%>%summary()
``` 
These are the summary statistics obtained for the selected numeric variables without grouping by a categorical. This is helpful because it provides a general overview of our data and serves as a point of comparison for when we begin the grouping process.

Now that we have a general overview of our numerics, let's further investigate by generating summary statistics for wind speed after grouping by tropical storm category.

```{r}
Final%>%group_by(category)%>%summarize(wind_min=min(wind), max_wind=max(wind), mean_wind=mean(wind), med_wind=median(wind), sd_wind=sd(wind), IQR_wind=IQR(wind))
```

As you may have already guessed, wind speed increases as tropical storm category increases.  

Instead of grouping by category, let's examine how wind speed differs when it's grouped by year.

```{r}
head(Final%>%group_by(year)%>%summarize(wind_min=min(wind), max_wind=max(wind), mean_wind=mean(wind), med_wind=median(wind), sd_wind=sd(wind), IQR_wind=IQR(wind)))
```

Here, we can see the breakdown for wind speed for every year. Something to note is the difference between the maximum wind speed and the average wind speed for a particular year. For example, let's have a look at year 2002. How can it be such that the max wind speed is significantly greater that the recorded average? As we will later find out, this suggests that that there are a far greater number of -1, 0, and 1 category storms for any given year. Thus, when it comes to average speed, the intense winds from higher category storms cannot outweigh lower category wind volume.

```{r}
Final%>%filter(year==2005)%>%group_by(status)%>%summarize(min_press=min(pressure), max_press=max(pressure), mean_press=mean(pressure), med_press=median(pressure), sd_wind=sd(pressure), IQR_press=IQR(pressure))
```

Earlier, we learned that 2005 had the greatest number of category 5 hurricanes. Grouping by status here gives us an overall breakdown of the numeric variable 'pressure' for the same year. Although this information does not pertain specifically to category 5 hurricanes since we grouped by status (which will include lower categories of the same classification), it is still useful because we can see the differences in pressure for storms occurring in 2005. It appears that pressure is negatively correlated with wind speed considering that hurricanes have a lower pressure than tropical storms, but produce much faster wind speeds. 

```{r}
Final%>%filter(between(year, 2005, 2009))%>%group_by(storm_cat)%>%summarize(min_birth=min(births), max_birth=max(births), mean_birth=mean(births), med_birth=median(births), sd_birth=sd(births), IQR_birth=IQR(births))
```

Above, we have grouped births by the categorical variable we created earlier: storm_cat. We can see from the output that during the five year period (2005-2009), the average number of births does not fluctuate significantly based on storm strength. 

```{r}
head(Final%>%group_by(year)%>%summarize(mean_birth=mean(births))%>%arrange(desc(mean_birth)))
```

Once again, we are assessing the numeric variable 'births', but have grouped by year instead of our categorical variable 'storm_cat'. It appears that 2007 had the highest number of mean births over the fourteen year period. 

Given that we just learned year 2007 had the highest number of mean births per day, it would be interesting to see if this year also reported a lower number of weaker storms. 

```{r}
Final%>%filter(year=="2007")%>%group_by(storm_cat)%>%summarize(n_s=n_distinct(name))
```

It appears that there were far more weak storms than medium/strong storms in 2007. Given this information, is it possible that storm intensity had an impact on the number of births in 2007?

All of the above examples involved grouping a single numerical variable to a single categorical variable. What if we're interested in grouping a single numerical variable to multiple categorical variables?

```{r}
head(Final%>%group_by(category, status)%>%summarize(mean_births=mean(births))%>%arrange(desc(mean_births)))
```

Here, we have grouped births by status and storm category. This is not suggestive of the idea that birth rate is at all impacted by storm strength considering that hurricanes are among the top average birth rates from 2000-2014.

Grouping several numeric variables to several categorical variables. 

```{r}
Final%>%group_by(category, storm_cat)%>%select(wind, pressure, category, storm_cat)%>%summarize_if(is.numeric, mean, na.rm=T)
```
 
Now that we have summarized a large portion of our data, we can generate a correlation matrix to better illustrate the relationship between our numeric variables.

```{r}
Final%>%na.omit()%>%select(births, wind, pressure, ts_diameter, hu_diameter)%>%cor()
```

As we previously inferred, the correlation matrix reveals that wind and pressure are strongly negatively correlated with each other. As wind speed increases, pressure decreases and vice versa. It also appears that wind is moderately positively correlated with ts_diamter and hu_diameter, meaning that pressure must be negatively correlated with these variables.

To better illustrate the relationship between wind speed and pressure, let's graph the mean wind speed (knots) against the mean wind pressure (millibars) by category over the fourteen year period.

```{r}
Final%>%group_by(category)%>%summarize(mean_wind=mean(wind), mean_pressure=mean(pressure))%>%ggplot(aes(mean_wind, mean_pressure, color=category))+geom_point(size=1.5)+ggtitle("Mean Wind Speed vs. Pressure by Storm Category from 2000-2014")+xlab("Wind Speed (knots)")+ylab("Pressure (millibars)")+theme_dark()
```

We can see that the weakest storms are found in the upper left hand side of the graph. These stroms have very low wind speeds but have the highest average pressure. Next, we can see that storms we categorized as medium (i.e. those ranging in category from 2 to 3) have greater wind speeds, but lower pressure averages than the weak storms. Finally, the category 4 and 5 storms found in the lower right hand corner have the highest average wind speed, but lowest average pressure. 

From the information provided in the correlation matrix, this graph can also be used as a template for interpreting the trends between other numeric variables. Wind speed is negative correlated with pressure, but positively correlated with two other variables: ts_diameter and hu_diameter. Although the relationship would not be nearly as linear if graphed, it would be expected for ts/hu_diamter to increase as category strength increases and decrease as category strength decreases. 

Let's create a graph that accounts for the average number of births by storm category over the fourteen year period. 

```{r}

Final%>%group_by(storm_cat,Storm_Month, category)%>%summarize(Births=mean(births))%>%ggplot(aes(Storm_Month, Births, fill=category))+geom_bar(stat="summary", fun.y="mean")+theme_gray()+scale_x_continuous(breaks=seq(1,12,1))+ggtitle("Mean Births by Storm Month and Category 2000-2014")+scale_fill_brewer(palette = 9)
```

We can see that birth frequency does not fluctuate significantly by category strength. For any category during any month over the fourteen year period, the average number of births per day is approximately 10,000. We can also see that category 5 storms only occurred between July - October. In terms of storm type, months 7-11 have the greatest storm variability. This likely represents the transition of a single storm from one category to the next as it becomes stronger with time. Lastly, we can see that no tropical storms were observed during February and March for any year from 2000-2014.

As was mentioned earlier, the frequency of a particular storm category can skew the average of a particular numeric if not carefully grouped prior to the creation of a summary statistic. This graph is an excellent example of why grouping by certain categoricals can more accurately define the parameter of interest.

PCA

```{r}
Final%>%na.omit()%>%select(wind, births, pressure, ts_diameter, hu_diameter)%>%scale%>%prcomp()->myPr
summary(myPr)
```

```{r}
Final%>%na.omit()%>%select(pressure, wind, births, hu_diameter, ts_diameter)%>%scale%>%princomp()->pca_Final
eigen<-pca_Final$sdev^2
varprop<-round(eigen/sum(eigen),2)
ggplot()+geom_bar(aes(y=varprop, x=1:5), stat="identity")+xlab("")+geom_path(aes(y=varprop, x=1:5))+geom_text(aes(x=1:5, y=varprop, label=round(varprop, 2)), vjust=1, col="white", size=5)+scale_y_continuous(breaks=seq(0, .6,.2), labels=scales::percent)+scale_x_continuous(breaks=1:10)
```

This screeplot suggests that we should include the first two principal components since together they can explain over eighty percent of the total variation in the 'Final' dataset. From the earlier summary, we can also see that the first two principal components are the only components with scores greater than one. Thus, using these two components to explain the majority of the observed variance in the dataset is a good approach.

```{r}
pca_Final$loadings[1:5, 1:2]%>%as.data.frame%>%rownames_to_column%>%ggplot()+geom_hline(aes(yintercept=0),lty=2)+geom_vline(aes(xintercept=0), lty=2)+ylab("PC2")+xlab("PC1")+geom_segment(aes(x=0, y=0, xend= Comp.1, yend=Comp.2), arrow=arrow(), col="red")+geom_label(aes(x=Comp.1*.8, y=Comp.2*.8, label=rowname))
```

These are the eigenvectors for each of the numerical variables within the modified dataset 'Final'. An eigenvector is simply the direction of the eigenvalue, or principal component. It defines the path along which you can expect to observe the greatest amount of variance for any particular cluster of data points. It provides the maximum orthogonal distance for any point along the line traveling through the cluster. Here, we can see that for closely related clusters (ts_diameter, hu_diameter, and wind), the eigenvectors are close together, opposite to that of the eigenvector for pressure, and orthogonal to the eigenvector for births. From this graph,it appears that PC1 reliably separates pressure from all eigenvectors on the left. PC2 reliably separates births from all other eignevectors. PC1, however, cannot reliably separate ts_diamter, hu_diameter, or wind from each other. More information would need to be included in order to separate these variables from one another.

Final Thoughts

Although I attempted to look at the relationship between storm intensity and birth rate, any of the summary statistics or graphical illustrations provided cannot possibly support my original hypothesis. There are simply too many confounding variables that may have contributed to any of the observed declines in birth rates reported in this project. In fact, I don't believe that my original hypothesis is even valid given the parameters of the datasets that I've chosen to work with. Tropical storms are mostly localized to a specific region of the United States (Gulf Coast/Southeasern shores) wherease birth data is collected for all states. Also, storms cannot delay the onset of delivery. If a woman is scheduled to give birth, it's not something that will be postponed because of the weather. With that being said, I enjoyed working on this project. I gained valuable experience learning how to peruse large datasets and employ functions to extract specific information. I now feel more confident in my ability to apply what I have learned in this course to real world scenarios. 
